---
title: "BMIN503/EPID600 Final Project"
author: "My Identity is a 'Secret'"
output: 
  html_document:
    keep_md: true
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
memory.limit(size=56000)
```  
***
### Overview

I want to look at if/how marital status impacts survival across different genders and cancer  sites. I know in general epi literature, there is a tendency to see that marriage typically is a stronger predictor of survival in men than in women depending on the disease.I plan to further stratify by race.


### Introduction 

It has been found across social epidemiological studies that marital status can be a predictor of survival outcomes for various diseases. Sometimes it is found that there is a  stronger association between being married and longer survival for men than for women. I  intend to investigate whether or not this pattern holds true for those diagnosed with various forms of cancer in the US and Puerto Rico, utilizing the SEER dataset. I would also like to further explore the effects of race on these potential associations. Epidemiology itself is an interdisciplinary field, and this project will also utilize the fields of social epidemiology, oncology, and biostatistics.

### Methods

I am using the SEER database below (over 8 million observations so far, so I can't upload it):
Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov) SEER*Stat Database: Incidence - SEER 18 Regs Research Data + Hurricane Katrina Impacted Louisiana Cases, Nov 2018 Sub (1975-2016 varying) - Linked To County Attributes - Total U.S., 1969-2017 Counties,National Cancer Institute, DCCPS, Surveillance Research Program, released April 2019, based on the November 2018 submission. 
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
#load raw dataset
seer <- read.csv("C:/Users/kmroh/Documents/BMI Cert/Fall 2020/BMIN503/dataset1.csv", header = T)
#view column names and rename variables of interest
colnames(seer)
seer <- seer %>%
  rename(patid = Patient.ID, 
         sex = Sex, 
         year.dx = Year.of.diagnosis,
         mstatus = Marital.status.at.diagnosis,
         race = Race.recode..White..Black..Other.,
         surv.months = Survival.months,
         cod.site = COD.to.site.recode,
         cod = SEER.cause.specific.death.classification,
         vital.status = Vital.status.recode..study.cutoff.used.,
         yob = Year.of.birth,
         prim.site = Primary.Site...labeled) %>%
  select(-year.dx,-yob)
seer$surv.months <- as.integer(seer$surv.months)
#see possible marital statuses
table(seer$mstatus)
#remove entries which are not married, partnered, or single at time of diagnosis
seer <- seer %>%
  filter(mstatus == "Married (including common law)" |
         mstatus == "Unmarried or Domestic Partner" |
         mstatus == "Single (never married)")
#recode to those who are partnered or not at the time of diagnosis
seer$part <- ifelse(seer$mstatus != "Single (never married)",
                    1,
                    0)
seer$part <- factor(seer$part, labels = c("Single","Partnered"))
#recode race variable
table(seer$race)
seer$race <- ifelse(seer$race == "Other (American Indian/AK Native, Asian/Pacific Islander)",
                    "Other", ifelse(seer$race == "White",
                                    "White", ifelse(seer$race == "Black",
                                                    "Black", ifelse(seer$race == "Unknown",
                                                                    "Unknown", 0))))
#remove those of unknown race 
seer <- seer %>%
  filter(race != "Unknown")
#remove those whose cause of death is unknown, or whose tumor is not the first
table(seer$cod)
seer <- seer %>%
  filter(cod == "Alive or dead of other cause Dead" |
         cod == "Dead (attributable to this cancer dx)")
#remove observations with unknown primary site and those that were alive at the study close
seer <- seer %>%
  filter(prim.site != "C80.9-Unknown primary site" |
           cod != "Alive")
#per classmate's feedback I'm setting a cap for survival months at 120 (10 years)
seer$surv.months <- ifelse(seer$surv.months >= 120, 120, seer$surv.months)
#create group average of survival month by part variable
seer <- seer %>%
  group_by(part) %>%
  mutate(avg.pstatus = mean(surv.months, na.rm = T))

#create group average of survival month by sex
seer <- seer %>%
  ungroup() %>%
  group_by(sex) %>%
  mutate(avg.sex = mean(surv.months, na.rm = T))

#create group average of survival month by mar and sex variables
seer <- seer %>%
  ungroup() %>%
  group_by(part,sex) %>%
  mutate(avg.part.sex = mean(surv.months, na.rm = T))

#recode primary site variable to body category
seer$c.code <- str_extract(seer$prim.site, "C[0-9][0-9]")
seer$c.code <- as.numeric(str_extract(seer$c.code, "[0-9][0-9]"))
seer$group <- ifelse(seer$c.code <= 14, "Lip, Oral Cavity and Pharynx",
 ifelse(seer$c.code <= 26,  "Digestive Organs",
 ifelse(seer$c.code <= 30,  NA,
 ifelse(seer$c.code <= 39,  "Respiratory and Intrathoracic Organs",
 ifelse(seer$c.code <= 41,  "Bone and Articular Cartilage",
 ifelse(seer$c.code == 42,  NA,
 ifelse(seer$c.code <= 44,  "Skin",
 ifelse(seer$c.code <= 49,  "Mesothelial and Soft Tissue",
 ifelse(seer$c.code == 50,  "Breast",
 ifelse(seer$c.code <= 58,  "Genital Organs",
 ifelse(seer$c.code == 59,  NA,
 ifelse(seer$c.code <= 63,  "Genital Organs",
 ifelse(seer$c.code <= 68,  "Urinary Tract",
 ifelse(seer$c.code <= 72,  "Eye, Brain or Central Nervous System",
 ifelse(seer$c.code <= 75,  "Thyroid & Endocrine Glands",
 ifelse(seer$c.code <= 80,  "Ill-Defined Sites",
 ifelse(seer$c.code <= 96,  "Lymphoid, Hematopoietic and Related Tissue", NA)))))))))))))))))
seer <- seer %>%
  ungroup() %>%
  group_by(part,sex,race) %>%
  mutate(avg.race = mean(surv.months, na.rm = T)) %>%
  ungroup() %>%
  group_by(part,sex,group) %>%
  mutate(avg.grp = mean(surv.months, na.rm = T)) %>%
  na.omit()

#create groups for pairwise ttests and anovas (there was probably a more elegant way to do this but this worked)
seer$psgroup[seer$sex == "Female" & seer$part == "Partnered"] <- "Part.Fem"
seer$psgroup[seer$sex == "Female" & seer$part == "Single"] <- "Sing.Fem"
seer$psgroup[seer$sex == "Male" & seer$part == "Partnered"] <- "Part.Male"
seer$psgroup[seer$sex == "Male" & seer$part == "Single"] <- "Sing.Male"
table(seer$psgroup)

seer$psrgroup[seer$sex == "Female" & seer$part == "Single" & seer$race == "White"] <- "White.Sing.Fem"
seer$psrgroup[seer$sex == "Female" & seer$part == "Single" & seer$race == "Black"] <- "Black.Sing.Fem"
seer$psrgroup[seer$sex == "Female" & seer$part == "Single" & seer$race == "Other"] <- "Oth.Sing.Fem"
seer$psrgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$race == "White"] <- "White.Part.Fem"
seer$psrgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$race == "Black"] <- "Black.Part.Fem"
seer$psrgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$race == "Other"] <- "Oth.Part.Fem"
seer$psrgroup[seer$sex == "Male" & seer$part == "Single" & seer$race == "White"] <- "White.Sing.Male"
seer$psrgroup[seer$sex == "Male" & seer$part == "Single" & seer$race == "Black"] <- "Black.Sing.Male"
seer$psrgroup[seer$sex == "Male" & seer$part == "Single" & seer$race == "Other"] <- "Oth.Sing.Male"
seer$psrgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$race == "White"] <- "White.Part.Male"
seer$psrgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$race == "Black"] <- "White.Part.Male"
seer$psrgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$race == "Other"] <- "Oth.Part.Male"
table(seer$psrgroup)

seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Bone and Articular Cartilage"] <- "Bone.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Breast"] <- "Breast.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Digestive Organs"] <- "Dig.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Eye, Brain or Central Nervous System"] <- "Eye.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Genital Organs"] <- "Gen.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Ill-Defined Sites"] <- "Ill.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Lip, Oral Cavity and Pharynx"] <- "Lip.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Mesothelial and Soft Tissue"] <- "Meso.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Respiratory and Intrathoracic Organs"] <- "Resp.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Skin"] <- "Skin.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Thyroid & Endocrine Glands"] <- "Thy.Sing.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Single" & seer$group == "Urinary Tract"] <- "Uri.Sing.Fem"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Bone and Articular Cartilage"] <- "Bone.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Breast"] <- "Breast.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Digestive Organs"] <- "Dig.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Eye, Brain or Central Nervous System"] <- "Eye.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Genital Organs"] <- "Gen.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Ill-Defined Sites"] <- "Ill.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Lip, Oral Cavity and Pharynx"] <- "Lip.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Mesothelial and Soft Tissue"] <- "Meso.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Respiratory and Intrathoracic Organs"] <- "Resp.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Skin"] <- "Skin.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Thyroid & Endocrine Glands"] <- "Thy.Sing.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Single" & seer$group == "Urinary Tract"] <- "Uri.Sing.Male"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Bone and Articular Cartilage"] <- "Bone.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Breast"] <- "Breast.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Digestive Organs"] <- "Dig.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Eye, Brain or Central Nervous System"] <- "Eye.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Genital Organs"] <- "Gen.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Ill-Defined Sites"] <- "Ill.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Lip, Oral Cavity and Pharynx"] <- "Lip.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Mesothelial and Soft Tissue"] <- "Meso.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Respiratory and Intrathoracic Organs"] <- "Resp.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Skin"] <- "Skin.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Thyroid & Endocrine Glands"] <- "Thy.Part.Fem"
seer$psbgroup[seer$sex == "Female" & seer$part == "Partnered" & seer$group == "Urinary Tract"] <- "Uri.Part.Fem"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Bone and Articular Cartilage"] <- "Bone.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Breast"] <- "Breast.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Digestive Organs"] <- "Dig.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Eye, Brain or Central Nervous System"] <- "Eye.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Genital Organs"] <- "Eye.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Ill-Defined Sites"] <- "Ill.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Lip, Oral Cavity and Pharynx"] <- "Lip.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Mesothelial and Soft Tissue"] <- "Meso.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Respiratory and Intrathoracic Organs"] <- "Resp.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Skin"] <- "Skin.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Thyroid & Endocrine Glands"] <- "Thy.Part.Male"
seer$psbgroup[seer$sex == "Male" & seer$part == "Partnered" & seer$group == "Urinary Tract"] <- "Uri.Part.Male"

```

### Results

Overall, both female sex and being partnered are associated with prolonged survival in cancer patients with statistically significant p-values of less than 0.05 in all ANOVAs and t-tests performed. The one caveat to the statistical significance of these results is that survival is generally only prolongued by a matter of a few months; while this may be emotionally meaningful for the loved ones of cancer patients, I also think that this could also mean prolongued suffering for terminally ill patients. Further, although the cleaned dataset has ~1.7 million observations, it is possible that that means the dataset is overpowered to detect significance. There are a few exceptions to the above stated pattern when subsetting the dataset by cancer sites. Specifically, bone and articular cartiledge patients survive longer when partnered, however this tendency is not statistically significant and there is virtually no difference between sexes; skin cancer patients are statistically differentiated between groups, except for single males and partnered females; mesothial single males are not statistically differentiated from females, whether single or partnered; breast patients are significantly differentiated by sex but not by Relationship status; partnered genital patients survive longer than single patients, however in this group men survive longer than women and these findings are statistically significant -- this pattern is also true for urinary tract patients; single male and female thyroid patients and ill-defined sites patients are not statistically differentiated.
```{r, warning=FALSE, message=FALSE}
library(rstatix)
library(ggpubr)
library(survival)
library(survminer)
library(ggrepel)

plots <- function(xvar, xname, gtitle, gavg){
  
  p <- ggplot(seer, aes(x=as.factor(xvar), y=surv.months)) +
  geom_violin(aes(middle = mean(surv.months))) +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black")) + 
  labs(title = gtitle, 
       x = xname, 
       y = "Survival Months") +
  stat_summary(fun.y=mean, colour="yellow") +
  geom_text(aes(label = round(gavg,2), y = gavg))
  print(p)
  #t-test for significance
  print(t.test(data = seer, surv.months~xvar))
  
  #kaplan-meier
  km_fit <- surv_fit(Surv(surv.months) ~ xvar, data=seer)
  km <- ggsurvplot(km_fit, data = seer, pval = T, conf.int = T, title = paste("Kaplan-Meier Survival Curve \nby ", xname),surv.median.line = "hv", palette = c("purple", "green"))
  geom_label_repel()
  print(km)
  #cox proportional hazards regression
  cox <- coxph(Surv(surv.months) ~ as.vector(unlist(xvar)), data = seer)
  print(summary(cox))
}

#plot survival months by recoded Relationship status
plots(seer$part, 
      "Relationship Status", 
      "Survival Months by Relationship Status at Diagnosis \nAll Sexes", 
      seer$avg.pstatus)

#plot survival months by sex
plots(seer$sex, 
      "Sex", 
      "Survival Months by Sex", 
      seer$avg.sex)

#plot survival months by relationship status and sex
  ggplot(seer, aes(x=as.factor(part), y=surv.months)) +
  geom_violin(aes(color = sex, middle = mean(surv.months))) +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position = "none") +
  labs(title = "Survival Months by Relationship Status at Diagnosis \nand Sex", 
       x = "Relationship status", 
       y = "Survival Months") +
  stat_summary(fun.y=mean, colour="yellow") +
  geom_text(aes(label = round(avg.part.sex,2), y = avg.part.sex)) +
  facet_wrap(~sex)
  #paired t-test for significance
  print(pairwise.t.test(seer$surv.months, seer$psgroup, p.adjust.method = "bonferroni"))
  #kaplan-meier
  km_fit <- surv_fit(Surv(surv.months) ~ part + sex, data=seer)
  gg <- ggsurvplot(km_fit, data = seer, pval = T, conf.int = T, title = paste("Kaplan-Meier Survival Curves \nby Relationship Status at Diagnosis and Sex"), legend.labs = c("Single","Single","Partnered","Partnered"),surv.median.line = "hv", palette = c("purple", "green")) 
  print(gg$plot + facet_wrap(~sex))
  
  #cox proportional hazards regression
  cox <- coxph(Surv(surv.months) ~ part + sex, data = seer)
  print(summary(cox))

#plot survival months by part, sex, and race
ggplot(seer, aes(x=as.factor(part), y=surv.months)) +
  geom_violin(aes(color = sex, middle = mean(surv.months))) +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position = "none") +
  labs(title = "Survival Months by Relationship Status at Diagnosis, \nSex, and Race", 
       x = "Relationship status", 
       y = "Survival Months") +
  stat_summary(fun.y=mean, colour="yellow") +
  facet_grid(~ race + sex) +
  geom_text(aes(label = round(avg.race,2), y = avg.race))
#pairwise t-test and anova based on partner status, sex and race
summary(aov(surv.months ~ psrgroup, data = seer))
pairwise.t.test(seer$surv.months, seer$psrgroup, p.adjust.method = "bonferroni")
#kaplan-meier
km_fit <- survfit(Surv(surv.months) ~ part + sex + race, data=seer)
gg <- ggsurvplot(km_fit, data = seer, pval = T, conf.int = T, title = paste("Kaplan-Meier Survival Curves \nby Sex, Race, and Relationship Status at Diagnosis"), legend.labs = c("Single","Single","Single","Single","Single","Single","Partnered","Partnered","Partnered","Partnered","Partnered","Partnered"),surv.median.line = "hv", palette = c("purple", "green"))
gg$plot + facet_wrap(~race + sex)
#cox proportional hazards regression
cox <- coxph(Surv(surv.months) ~ part + sex + race, data = seer)
summary(cox)

#plot survival months by part, sex and group

plots.group <- function(gname){
p <- ggplot(subset(seer, group %in% c(gname)), aes(x=as.factor(part), y=surv.months)) +
  geom_violin(aes(color = sex, middle = mean(surv.months))) +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position = "none") +
  labs(title = paste("Survival Months by Relationship Status at Diagnosis \nand Sex \nGroup: ", gname), 
       x = "Relationship status", 
       y = "Survival Months")  +
  stat_summary(fun.y=mean, colour="yellow") +
  geom_text(aes(label = round(avg.grp,2), y = avg.grp)) +
  facet_wrap(~sex)
print(p)
#pairwise t-test based on partner status, sex and cancer site
seer1 <- subset(seer, group %in% c(gname))
print(pairwise.t.test(seer1$surv.months, seer1$psbgroup, p.adjust.method = "bonferroni"))
#kaplan-meier
  km_fit <- surv_fit(Surv(surv.months) ~ part + sex, data=seer1)
  km <- ggsurvplot(km_fit, data = seer1, pval = T, conf.int = T, title = paste("Kaplan-Meier survival Curves \nGroup: ",gname), legend.labs = c("Single","Single","Partnered","Partnered"),surv.median.line = "hv", palette = c("purple", "green"))
  km <- km$plot + facet_wrap(~ sex)
  print(km)
  #cox proportional hazards regression
  cox <- coxph(Surv(surv.months) ~ part + sex, data = seer1)
  print(summary(cox))
}

plots.group("Lip, Oral Cavity and Pharynx")
plots.group("Digestive Organs")
plots.group("Respiratory and Intrathoracic Organs")
plots.group("Bone and Articular Cartilage")
plots.group("Skin")
plots.group("Mesothelial and Soft Tissue")
plots.group("Breast")
plots.group("Genital Organs")
plots.group("Urinary Tract")
plots.group("Eye, Brain or Central Nervous System")
plots.group("Thyroid & Endocrine Glands")
plots.group("Ill-Defined Sites")

```